\TITLE

### 2021-01-27 draft/response-3.tex

\documentclass[letterpaper, inpress]{jds} % use this for production
%\documentclass[a4paper, review]{jds}      % use this for review
\usepackage{lineno}
\linenumbers

\usepackage{subcaption}
\usepackage{calc}
\usepackage{comment}
\usepackage{todonotes}
\setuptodonotes{inline}

% color management

\usepackage{color}
\usepackage{tcolorbox}

% table management

%\usepackage[table]{xcolor}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating, multirow}
\usepackage{colortbl}

% Avoid non breaking words

\renewcommand{\_}{%
    \textunderscore\hspace{0pt}%
}
\setlength{\emergencystretch}{3em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% the following edits should be done by Journal typesetters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{page}{1}            % set the first page number
\jdsmonth{January}              % month
\jdsyear{2021}                  % year
\jdsvolume{xx}                  % volume number
\jdsissue{xx}                   % issue number
\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
\jdsreceived{Jun 30, 2020}       % optional: comment it out if no received date
\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
%% manually set running header for a shorter list of authors if needed
% \shortauthors{A Author A, et al.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% edits by authors are given below
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{booktabs}

\newtheorem{requirement}{Requirement}
\newtheorem{question}{Question}

\newcommand{\TITLE}{AICov: An Integrative Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates}

\newcommand{\ABSTRACT}{ The COVID-19 (COrona VIrus Disease 2019)
  pandemic has had profound global consequences on health, economic,
  social, behavioral, and almost every major aspect of human
  life. Therefore, it is of great importance to model COVID-19 and
  other pandemics in terms of the broader social contexts in which
  they take place.  We present the architecture of an artificial
  intelligence enhanced COVID-19 analysis (in short AICov) , which
  provides an integrative deep learning framework for COVID-19
  forecasting with population covariates, some of which may serve as
  putative risk factors.  We have integrated multiple different
  strategies into AICov, including the ability to use deep learning
  strategies based on Long Short-Term Memory (LSTM) and event
  modeling.  To demonstrate our approach, we have introduced a
  framework that integrates population covariates from multiple
  sources. Thus, AICov not only includes data on COVID-19 cases and
  deaths but, more importantly, the population's socioeconomic,
  health, and behavioral risk factors at their specific locations.
  The compiled data are fed into AICov, and thus we obtain improved
  prediction by the integration of the data to our model as compared
  to one that only uses case and death data. As we use deep learning
  our models adapt over time while learning the model from past data.}

\title[AICov: Deep
    Learning Framework for COVID-19 Forecasting with Population
    Covariates]{\TITLE}

\author[1]{Geoffrey C. Fox}
\author[1]{Gregor von Laszewski\thanks{Corresponding author. Email: laszewski@gmail.com}}
\author[1]{Fugang Wang}
\author[2,3]{Saumyadipta Pyne}

\affil[1]{Digital Science Center, Indiana University, Bloomington Indiana, USA}
\affil[2]{Public Health Dynamics Laboratory, and Department of Biostatistics,
  University of Pittsburgh, Pittsburgh, Pennsylvania, USA}
\affil[3]{Health Analytics Network, Pennsylvania, USA}

\begin{document}

\input{response-v3-content}

\maketitle

\input{response-v3-content}

\begin{abstract}
  \ABSTRACT
\end{abstract}

\begin{keywords} % alphabetical; excluding anything in the title already
  Cloudmesh;
  Comorbidities;
  Prediction;
  Risk factors.
\end{keywords}

\section{Introduction}

The COVID-19 pandemic has caused an enormous health and humanitarian
crisis worldwide. It is unlike any other single phenomenon that has
occurred in modern history since the end of World War II.  The
pandemic’s effects have spanned over a range that is so vast over
space, and yet so condensed over time, that the dual blows of
intensity and rapidity have exposed myriad systemic vulnerabilities in
many societies around the world.

The first known case was traced back to 17 November 2019. Since the
emergence of a cluster of cases in Wuhan, China, on 31 December 2019,
COVID-19 has spread rapidly worldwide. Just six months later, as of 23
June 2020, there are 8,993,659 COVID-19 cases and 469,587 deaths
globally. On 30 January 2020, the World Health Organization (WHO)
\citep{www-who} declared COVID-19 to be a Public Health Emergency of
International Concern and subsequently, on 11 March 2020, a pandemic.

In the United States, the first case of COVID-19 was confirmed on 20
January 2020 in Washington State, and as of 23 June 2020, there were
over 2.3 million confirmed cases and 121,167 deaths
\citep{www-nyt-map}.  Although initial efforts to reduce the spread
took place, recent data shows a resurgence at an increased scale. In
the absence of a vaccine or treatment to effectively combat the
disease, non-pharmaceutical policy interventions such as social
distancing and lockdowns are recommended by health experts to prevent further transmission. To gain insights into the possible impact of
such measures on COVID-19 outcomes, we depend upon the ability to
accurately forecast the spread of reported cases and confirmed deaths
and recoveries. Naturally, the accuracy of forecasting relies on the
availability of current, reliable data and historical data to
determine estimates of uncertainty.

During outbreaks of epidemics, initially data on cases and deaths
could be scarce, and the quality of data annotation, validation, and
aggregation might be uncertain. For instance, changes in clinical data
entry, such as the addition of a new category clinically diagnosed to
the existing lab-confirmed category, may likely have been reflected in
the reporting \citep{Petropoulos2020-hh}. Initial fears about COVID-19
among the general population with regards to the trajectory of the
pandemic could also affect administrative reporting. Diverse media
sources and differences in local and federal policies may add to the
general uncertainty about disease progression. Nevertheless, a
data-driven approach to forecasting can offer valuable insights into
the disease dynamics, and thereby an ability to objectively plan for
the near future.

Unlike earlier global outbreaks, during COVID-19, the current
worldwide digital ecosystem allows for real-time data collection,
which in turn is used by artificial intelligence (AI) and deep
learning systems to understand healthcare trends, model risk
associations, and predict outcomes \citep{Ting2020-lw}.  In addition to
traditional public health surveillance strategies available to most
countries, a variety of static and dynamic data types may be
integrated to model the scale and dynamics of this pandemic
\citep{Pyne2015-ao}. Several organizations, including the Johns Hopkins University’s Center for Systems Science and Engineering, the New York
Times, the Atlantic’s COVID-19 tracking project, among others have
developed real-time tracking maps for following COVID-19 cases around
the world using data from the U.S.~Centers for Disease Control and
Prevention (CDC), WHO, and other international health
agencies. Notably, the CDC receives forecasts from several modeling
groups that use a wide variety of approaches ranging from SEIR
(Susceptible-Exposed-Infectious-Recovered) to Agent-based to Bayesian
models to understand the non-pharmaceutical policy interventions (or the lack thereof) to predict disease dynamics and its impact on human
lives \citep{www-cdc-modeling-forecast}.

For precise and contextually relevant modeling of COVID-19 dynamics
for a given population or community, we use data integration to
combine community-specific health and behavioral risk factors,
demographic and socioeconomic variables that are available for the corresponding county along with the cases and deaths data for COVID-19 while
leveraging the recently developed and increasingly popular deep
learning strategies to forecast COVID-19 dynamics using our
model. Towards this, we develop a parallel computing platform called
AICov standing for AI-driven Platform for COVID-19 to implement the
different components of our framework and obtain the needed resources
through multi-cloud interfaces in a robust manner.

In this study, while we incorporate information from community and
county-specific covariates to inform our forecasting model for each
metropolitan area, we want to avoid the so-called ecological fallacy
in drawing inferences about individual disease outcomes based on large
area-level aggregated risk factors. Our objective is also to
underscore the nuanced roles played by pre-existing socioeconomic and
other prevailing conditions of a given community that can act as
determinants of its health outcomes and possible disparities,
especially under such sudden stresses to its current systems as those
felt during any pandemic.  It is important to note that the underlying
models are learned directly from the data instead of being projected
by a theoretical epidemiological model that relies on simulation
populations. This allows quick adaptation and self-adaptation in
regards to changing behaviors and regional differences.

The paper is structured as follows. In Section \ref{sec:lstm-theory},
we provide a short overview of deep learning-based time series
forecasting that we use in AICov. In Section \ref{sec:arch}, we
outline our architecture that enables the user to conduct powerful
time series forecasting by integrating static and dynamic information
on population risk factors with time-series disease data. In Section
\ref{sec:data}, we describe the input data for AICov. Next, we
demonstrate our analysis involving the population risk factors and
identify those that contribute to our prediction accuracy. Finally, we
present our conclusion.

\section{LSTM}
\label{sec:lstm-theory}

This section presents a short overview of the recurrent neural network
algorithms used in this work.

% ------------------------------
\subsection{LSTM Background}
\label{sec:lstm-background}
% ------------------------------

In this study, we base our work on a traditional LSTM algorithm for our predictions
\citep{www-keras-lstm,Hochreiter1997-dk}. An LSTM is a Recurrent
Neural Network (RNN) \citep{Rumelhart1986-li} with feedback
connections allowing the use of data input sequences to predict data
output sequences. LSTMs have been applied to many application areas
from handwriting \citep{Graves2009-qb}, image, feature detection, and
time series prediction \citep{Schmidhuber2005-oy}. LSTM's have an
internal state that is used to prevent the vanishing gradient problem
\citep{Hochreiter1991-mp} during the training of RNNs.

Different variants of LSTM algorithms exist. Ours is based on an LSTM
cell depicted in Figure \ref{fig:lstm}.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.6\columnwidth]{images/lstm.pdf}
    \caption{LSTM Unit Diagram}
    \label{fig:lstm}
\end{figure}

It has an input gate, output
gate, and a forget gate. The cell is maintaining how a subsequent
value is calculated. This includes (a) how input values are
influencing the cell via the input gate, (b) how the forget gate
influences a memory value within the cell via the forget gate, and (c)
how the output gate influences the output activation of the LSTM
unit. A cell has a number of inputs and outputs that are
weighted. During a training step, these weights are learned to be
reused in a prediction process between input and output sequences.
The LSTM network will have a number of input values represented as  vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}. For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell. In our implementation the number for inputs and outputs can be adjusted. We repeat this prediction over all input sequences to create the output forecast sequences. Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.

\newcommand{\RH}{\mathbb{R}^{H}}
\newcommand{\RD}{\mathbb{R}^{D}}
\begin{align}
    f_t &= \sigma_g(W_f x_t + U_f h_{t-1} + b_f) \label{eq:first} \\
    i_t &= \sigma_g(W_i x_t + U_i h_{t-1} + b_i) \\
    o_t &= \sigma_g(W_o x_t + U_o h_{t-1} + b_o) \\
    \tilde{c}_t &= \sigma_c(W_c x_t + U_c h_{t-1} + b_c) \\
    c_t &= f_t \circ c_{t-1} +i_t \circ \tilde{c}_t  \\
    h_t &= o_t \circ \sigma_h(c_t) \label{eq:last}
\end{align}

We use in the euqations $\circ$ as the Hadamard product,
$\sigma_g$ is the sigmoid function for the gate,
$D$ the number of days as input features, and
$H$ the number of hidden layers.
Inputs and outputs are defined by 
$x_{t} \in \RD$ the input vector to the LSTM unit, and 
$h_{t}\in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. Other variables include
$f_t \in \RH$ is the forget gate's activation vector, 
$i_t \in \RH$ is the input/update gates activation vector, 
$o_{t}\in \RH$ output gate's activation vector, 
$\tilde{c}_{t}\in \RH$ the cell input activation vector
$c_{t}\in \RH$ the cell state vector,  
$\sigma_c$ is the cell activation function, 
$W\in \mathbb {R} ^{H\times D}$, $U\in \mathbb {R} ^{H\times H}$ and $b\in \RH$ being the weight matrices and bias vector parameters which are learned during training.
In particular we denote $W_{q}$ and $U_{q}$ contain the weights of the
input and recurrent connections, where the subscript denotes either
the input gate $i$, the output gate $o$, the forget gate $f$ or the
memory cell $c$. For a more detailed discussion about LSTM's we refer to \cite{lstm-explained}.

\subsection{Covariate Enhanced LSTM}

We often think of the laws of physics described by operators that
evolve the system given sufficient initial conditions and in this
language, we have shown how to represent Newton’s law operator by a
recurrent network \citep{Kadupitiya2020-zq}. We expect that the time
dependence of many complex systems: Covid pandemics, Southern
California earthquakes, traffic flow, security events can be described
by deep learning operators that both capture the dynamics and allow
predictions.

In this paper, we adopt this approach to describe the
time dependence of COVID-19 infection and fatality counts in different
regions. 
We intend to
process the two time-series (infections, fatalities/deaths) associated with
each city plus the covariates (fixed in time but dependent on region)
to build a model for the time evolution of the data. There is a
nontrivial architecture design choice as to the method of handling the
time-independent covariates. We could combine a recurrent network for
the time series with some fixed network for the covariates. However,
that does not fit with our concept of deriving an evolution operator
for COVID-19 data that naturally depends on the covariates for each
city. Thus instead, we feed the covariates into the recurrent network as input in addition to the days.

\section{AICov Architecture}
\label{sec:arch}

The motivation for creating this architecture is based on our
experience with deep learning toolkits such as Keras and
PyTorch. While these systems provide the necessary APIs to integrate
time series solutions, they target primarily more general user
communities. In particular, they do not consider existing data sets or
specific needs and analysis options to evaluate them for the purpose
of COVID-19 risk factor data integration. We describe below the
different requirements that motivate the design of our architecture
framework, AICov, and how we address each of these.

%\newcommand{\Solution}{{\bf Solution: }}
\newcommand{\Solution}{}

\begin{requirement}{\bf Ease of Use.} \normalfont
  A major design criterion for AICov is that the framework must be
  easy to use, and extensions can be made to allow usability of both
  the Application Programming Interfaces (API) as well as the user
  interface. \Solution To address these requirements, we are devising
  a specialized but easy to use time series API for COVID-19 that
  integrates common tasks such as automated filtering and
  normalization of the data.
\end{requirement}

\begin{requirement}{\bf Interactive Access.} \normalfont
  Due to the experimental nature of analyzing the data, AICov must
  support the exploration of the data in an interactive
  fashion. \Solution As many data scientists use Jupyter Notebooks to
  integrate their work interactively, AICov integrates such
  notebooks. Jupyter notebooks allow rapid modification of the
  analysis workflow through the ease of Python as programming
  language. Additionally, they enable us to formulate easily
  sophisticated scientific analytics workflows with the help of
  Python and Jupyter Notebooks.
\end{requirement}

\begin{requirement}{\bf Expandable API.} \normalfont
  Assuming that new models and other analysis algorithms will be
  developed over time, AICov must allow the integration of such
  models through programs and APIs. \Solution To address this 
  requirement, we are developing an
  abstract API for data, analysis, and metadata parameter
  adjustments. External services can be integrated with the help of
  REST (REpresentational State Transfer) services. The REST services
  are automatically generated from function specifications \citep{las-19-openapi-function}.
\end{requirement}

\begin{requirement}{\bf Flexible Data Source Integration.} \normalfont
  The integration of various sources of data is a key part of
  AICov. \Solution To achieve this goal, we provide a number of
  abstractions and data manipulation functions to extract needed data
  from established sources. Furthermore, the data can be combined, and
  additional data wrangling by external groups can be integrated.
  Through such abstractions, it is possible to replace, add, and
  correct data sets that are part of the analysis. This allows us to
  compare different results created from different data
  sets. Moreover, we include data selection directly in our framework
  to easily and quickly reuse them in a data mashup. An important
  example is that the data need not be static over the disease's
  progression but can be continuously updated. This also has a
  significant impact on our analytics algorithms.
\end{requirement}

\begin{requirement}{\bf Automated Update of the Analytics.} \normalfont
  As the data can change, previous analyses may have to be
  updated. \Solution Our forecast is not just a single script but
  includes a mechanism to register multiple workflows of continuously
  integrated (CI) analysis that are automatically rerun when new data
  are made available to the system. Previous results are
  maintained. Metadata with these runs can retrieve the versioned data
  sets used to calculate and store the versioned analytics workflow.
\end{requirement}

\begin{requirement}{\bf Flexible Model definition.} \normalfont
  We want to allow experimentation with various models in AICov that
  may be easily and flexibly definable as workflows. \Solution We give
  functional abstraction definitions that enables us to define the
  models ourselves or integrate third party models describing the
  spread of the disease.
\end{requirement}

\begin{requirement}{\bf Deep Learning Forecast.} \normalfont
  As the forecasting models depend on changing data that are made
  available daily, it is important for AICov to utilize the newly
  available data. Many models can be applied to this, including moving
  averages based models or models that are purely derived from deep
  learning. \Solution To address this requirement, AICov allows for
  the integration of both. Furthermore, in our deep learning
  framework, we can integrate an automated search for important risk
  factors, which is the focus of this study.
\end{requirement}

### 2021-01-27 final/vonLaszewski-covid.tex
add final from draft

It has an input gate, output
gate, and a forget gate. The cell is maintaining how a subsequent
value is calculated. This includes (a) how input values are
influencing the cell via the input gate, (b) how the forget gate
influences a memory value within the cell via the forget gate, and (c)
how the output gate influences the output activation of the LSTM
unit. A cell has a number of inputs and outputs that are
weighted. During a training step, these weights are learned to be
reused in a prediction process between input and output sequences.
The LSTM network will have a number of input values represented as vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}. For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell. In our implementation the number for inputs and outputs can be adjusted. We repeat this prediction over all input sequences to create the output forecast sequences. Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.  We use in the equations $\circ$ as the Hadamard product,
$\sigma_g$ is the sigmoid function for the gate.
With 
$D$ being the number of days as input features, and
$H$ being the number of hidden layers. 
The inputs and outputs are defined by  
$x_{t} \in \RD$ the input vector to the LSTM unit, and 
$h_{t} \in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. The equations are

>>>>

It has an input gate, output
gate, and a forget gate. The cell is maintaining how a subsequent
value is calculated. This includes (a) how input values are
influencing the cell via the input gate, (b) how the forget gate
influences a memory value within the cell via the forget gate, and (c)
how the output gate influences the output activation of the LSTM
unit. A cell has a number of inputs and outputs that are
weighted. During a training step, these weights are learned to be
reused in a prediction process between input and output sequences.
The LSTM network will have a number of input values represented as vector x and a number of output values in the last layer of the network predicting a number of future values \citep{lstm-explained}. For our experiments this includes the numbers of days we use as input to our prediction and the number of days we use as output of our prediction for a single cell. In our implementation the number for inputs and outputs can be adjusted. We repeat this prediction over all input sequences to create the output forecast sequences. Next we list the equations (\ref{eq:first}) - (\ref{eq:last}) used in LSTM.  We use in the equations $\circ$ as the Hadamard product,
$\sigma_g$ is the sigmoid function for the gate.
With 
$D$ being the number of days as input features and
$H$ being the number of hidden layers, the 
inputs and outputs are defined by  
$x_{t} \in \RD$ the input vector to the LSTM unit and 
$h_{t} \in \RH$ the hidden state vector e.g., the last of which is the output vector of the LSTM. In our case the output contains the days we predict based on the input features. The equations are
